Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-13 agujournaltemplate.tex
Implemented final comments

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". But, to date, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means). Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient Forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

    \begin{table}
    \caption{Data and corresponding units used as predictor variables for the Gradient Forest analysis.}
    \centering
    \resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
    \toprule
    Variable Name & Units  & Data source \\
    \midrule
    MEI v.2 & unitless & Climate index provided by NOAA \\
    AMO & unitless & Climate index provided by NSF NCAR \\
    Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
    Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
    Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
    SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
    \qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
    Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[NO_3]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[PO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[SiO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    \end{tabular}}
    \label{tab:DataSources}
    \end{table}

@@

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts}$b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts}$a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly chlorophyll \textit{a} measurements for the remainder of the year 2013 were markedly lower (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

@@

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
    \caption{Clustering analysis. (a) Hierarchical clustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal features. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. The dots are color-coded based on the cluster they belong to. Vector overlay of ENVFIT shows how environmental variables co-vary with community data. Length of vector indicates strength of relationship.}
    \label{fig:clustering}
    \end{figure}

@@

\subsection{AMO and upwelling-related variables drive community changes}
    % Figure 6
    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community. Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera). 
    Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
    \caption{Gradient Forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
    \label{fig:GF}
    \end{figure}

@@

    % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
    The temporal dynamics found in yearly mean genus richness, however, did not correspond directly to the observed clusters. Instead, there was a marked reduction from 1998 to 2010, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1. The same trend was evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlated with genus richness and exhibited a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
    We did not find a significant difference in diversity between the two clusters, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}). However, we did observe a significant difference in Pielou's index between the two clusters. The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data. This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$). As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

\subsection{Drivers of change}
    % P4: Introduce GF Analysis and discuss results
    To identify the variables that best predict changes in the phytoplankton community, we conducted a Gradient Forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts. Gradient Forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis revealed that the strongest predictor of change within the community was not an in situ measurement but rather the AMO climate index. The AMO is followed in importance by SST, MEI v.2 index, and nitrate concentration as predictive variables (see Figure \ref{fig:GF}).

    % P5: Climate impacts
    Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, showed contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions. The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming more frequently during the time period of 2006 to 2010, which was characterized by mostly positive AMO anomalies.

@@

    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which co-varied most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations varied more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibited some explanatory power in the Gradient Forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}). This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.

@@

    % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
    Using microscopy cell counts as a basis for our analyses allowed us to focus on changes in the microphytoplankton community. Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the Gradient Forest model was further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$). This does not represent a complete picture of the phytoplankton community of the Cariaco basin. The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories. With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained. Drastic declines in cell counts reflected shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.

>>>>

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". But, to our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means. Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a gradient forest analysis. Gradient forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the gradient forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

    \begin{table}
    \caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
    \centering
    \resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
    \toprule
    Variable Name & Units  & Data source \\
    \midrule
    MEI v.2 & unitless & Climate index provided by NOAA \\
    AMO & unitless & Climate index provided by NSF NCAR \\
    Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
    Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
    Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
    SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
    \qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
    Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[NO_3]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[PO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[SiO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    \end{tabular}}
    \label{tab:DataSources}
    \end{table}

@@

    % Figure 3
    Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons. The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters. Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts} $b$). The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters. In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts} $a$). The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards. Nevertheless, the monthly chlorophyll \textit{a} measurements for the remainder of the year 2013 were markedly lower (65\,\% of the mean over the entire time series for these months), which recovered in 2014. Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series. In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.

@@

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
    \caption{Clustering analysis. (a) Hierarchical clustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance. The community for each year is binned together, to smooth out seasonal features. (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level. The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom. (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed. Stress: 0.261, 2 dimensions, converging solution. The dots are color-coded based on the cluster they belong to. Vectors show how environmental variables co-vary with community data. Length of vector indicates strength of relationship.}
    \label{fig:clustering}
    \end{figure}

@@

\subsection{AMO and upwelling-related variables drive community changes}
    % Figure 6
    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity. The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community. Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$. The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms. However, two haptophyte and three dinoflagellate genera rank relatively high. For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera). 
    Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range. The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria. Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition. Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here. For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
    \caption{Gradient forest analysis of monthly community data at a genus level. (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances. (b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance. (c) Cumulative importance curves for the predictor variables. Black line shows cumulative importance across predictor range for the entire model and all genera. Dashed colored lines represent the responses of individual functional groups.}
    \label{fig:GF}
    \end{figure}

@@

    % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
    The temporal dynamics found in yearly mean genus richness, however, did not correspond directly to the observed clusters. Instead, there was a marked reduction from 1998 to 2010, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1. The same trend was evident, albeit to a lesser degree, in the other diversity metrics. Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlated with genus richness and exhibited a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
    We did not find a significant difference in diversity between the two clusters, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}). However, we did observe a significant difference in Pielou's index between the two clusters. The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeA{pinckney_phytoplankton_2015} based on HPLC pigment data. This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community. The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$). As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.

\subsection{Drivers of change}
    % P4: Introduce GF Analysis and discuss results
    To identify the variables that best predict changes in the phytoplankton community, we conducted a gradient forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts. Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis revealed that the strongest predictor of change within the community was not an in situ measurement but rather the AMO climate index. The AMO is followed in importance by SST, MEI v.2 index, and nitrate concentration as predictive variables (see Figure \ref{fig:GF}).

    % P5: Climate impacts
    Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, showed contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions. The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeA{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming more frequently during the time period of 2006 to 2010, which was characterized by mostly positive AMO anomalies.

@@

    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which co-varied most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations varied more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}). This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.

@@

    % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
    Using microscopy cell counts as a basis for our analyses allowed us to focus on changes in the microphytoplankton community. Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the gradient forest model was further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$). This does not represent a complete picture of the phytoplankton community of the Cariaco basin. The HPLC pigment data used by \citeA{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories. With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained. Drastic declines in cell counts reflected shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.
